Supplementary Online Material for Landform boundary effects on Holocene forager landscape use in arid South Australia

### 2017-02-01 analysis/paper/SOM.Rmd
edits from PH, PH and MS to adapt the paper for the current NT conditions.

Marjorie Sullivan    
The Australian National University, Canberra ACT 0200, Australia and 
Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia

@@

The decision to use a mobile GIS was based on four specific advantages over paper-based systems of the numerous benefits listed by Tripcevich (2004a; 2004b), Clarke (2004) and Wagtendonk and De Jeu (2007):

@@

3.	There were other digital datasets, including fence and road locations, previously surveyed areas and recorded sites, and terrain units that covered all of the survey area. Over 95% of the survey was covered by good quality recent air photography. Some of the thematic maps had been digitized, other data were able to be digitized and stored as shapefiles by the mining company, BHP Billiton. While paper maps can provide this information, and may be easier to read than a small screen in bright sun, a mobile GIS gives more immediate context awareness (Pascoe et al. 1999). In particular the mobile GIS enables a user to determine location relative to other features and automates calculations relating to these features such as distance, direction and area.

@@

Data collection with a mobile GIS enabled our survey team to collect the massive amount of data collected during survey fieldwork and substantially reduce transcription time and error (Fitzpatrick 2002; McPherron & Dibble 2003; Tripcevich 2004a; 2004b). Our experience is comparable to that of Döner and Yomralıoğlu (2008), who recorded the time spent on field data collection and transferring data using both mobile GIS and a paper and GPS system for a project mapping transport networks for rural settlements in Turkey. They observed that the most substantial time savings were gained during data transfer. The pattern they document is a three-fold increase in the rate of site recording with mobile GIS. The same general pattern of time saving was observed in our history of research at Olympic Dam. Our pre-2007 data was transferred from field notes to a computer in 8-10 days after the survey concluded, while our 2007-2009 data was transferred from the mobile GIS to the desktop computer in less than five minutes per fieldwork day.

The Olympic Dam sites fall into a narrow range of site types and stone artefacts within them are patterned in highly repeated ways that suited a systematic recording protocol. An important consideration in evaluating this use of mobile GIS and traditional methods is the regularity of the archaeological materials that were recorded. Projects that encounter high levels of variation, or cannot anticipate the kinds of data to be recorded, might benefit more from freeform data capture rather than the tightly controlled data entry fields that were an important advantage for this survey. Of course the value of more flexible data entry and subsequent problems for analysis would need to be balanced against the enhanced uniformity and accuracy of data when entry is tightly controlled and subject to automatic error checking and defined value limits.

As a final observation, it is noted that the disruption to fieldwork when the mobile GIS system fails may be underrepresented in the previous literature on mobile GIS in archaeology (McPherron & Dibble 2003; Tripcevich 2004b; but see McCaffrey et al. 2005). Mobile GIS is a complex and flexible setup that replaces many pieces of equipment (including tape measures, compass, notebooks) and if the handheld computer fails during a paperless survey then fieldworkers are limited in the tasks they can perform. This was a more serious concern in the early implementation of mobile GIS and more recent equipment and software are mostly reliable and robust. Breakdowns can never be completely avoided however and during the Olympic Dam survey their impacts were mitigated with multiple nightly data backups and keeping redundant functioning equipment on hand for immediate replacement of broken units, to minimize delays caused by waiting for repairs.  When the sites database was 'cleaned' at the end of the survey stage, serious errors (ie. fields containing data outside of a credible range or containing a mix of integers and non-integers, mostly likely due to accidents during data entry) and omissions (fields left blank or with data that is incomplete) were found to have occurred in fewer than 0.5% of the sites. Sites with problematic data were re-recorded and the unclean recorded were marked so they could be excluded from analysis.

@@

>>>>

Marjorie Sullivan    
The University of Sydney, Sydney NSW, Australia

@@

The decision to use a mobile GIS was based on four specific advantages over paper-based systems of the numerous benefits listed by Tripcevich (2004a; 2004b), Clarke (2004) and Wagtendonk and De Jeu (2007), and previously summarised for this project by Sullivan et al. (2014):

@@

3.	There were other digital datasets, including fence and road locations, previously surveyed areas and recorded sites, and terrain units that covered all of the survey area. Over 95% of the survey was covered by good quality recent air photography. Some of the thematic maps had been digitized, other data were able to be digitized and stored as shapefiles by the mining company's land managers. While paper maps can provide this information, and may be easier to read than a small screen in bright sun, a mobile GIS gives more immediate context awareness (Pascoe et al. 1999). In particular the mobile GIS enables a user to determine location relative to other features and automates calculations relating to these features such as distance, direction and area.

@@

Data collection with a mobile GIS enabled our survey team to collect the massive amount of data collected during survey fieldwork and substantially reduce transcription time and error (Fitzpatrick 2002; McPherron & Dibble 2003; Tripcevich 2004a; 2004b). Our experience is comparable to that of Döner and Yomralıoğlu (2008), who recorded the time spent on field data collection and transferring data using both mobile GIS and a paper and GPS system for a project mapping transport networks for rural settlements in Turkey. They observed that the most substantial time savings were gained during data transfer. The pattern they document is a three-fold increase in the rate of site recording with mobile GIS. The same general pattern of time saving was observed in our history of research in the Roxby dunefield. Our pre-2007 data was transferred from field notes to a computer in 8-10 days after the survey concluded, while our 2007-2009 data was transferred from the mobile GIS to the desktop computer in less than five minutes per fieldwork day.

The Roxby dunefield sites fall into a narrow range of site types and stone artefacts within them are patterned in highly repeated ways that suited a systematic recording protocol. An important consideration in evaluating this use of mobile GIS and traditional methods is the regularity of the archaeological materials that were recorded. Projects that encounter high levels of variation, or cannot anticipate the kinds of data to be recorded, might benefit more from freeform data capture rather than the tightly controlled data entry fields that were an important advantage for this survey. Of course the value of more flexible data entry and subsequent problems for analysis would need to be balanced against the enhanced uniformity and accuracy of data when entry is tightly controlled and subject to automatic error checking and defined value limits.

As a final observation, it is noted that the disruption to fieldwork when the mobile GIS system fails may be underrepresented in the previous literature on mobile GIS in archaeology (McPherron & Dibble 2003; Tripcevich 2004b; but see McCaffrey et al. 2005). Mobile GIS is a complex and flexible setup that replaces many pieces of equipment (including tape measures, compass, notebooks) and if the handheld computer fails during a paperless survey then fieldworkers are limited in the tasks they can perform. This was a more serious concern in the early implementation of mobile GIS and more recent equipment and software are mostly reliable and robust. Breakdowns can never be completely avoided however and during the field survey their impacts were mitigated with multiple nightly data backups and keeping redundant functioning equipment on hand for immediate replacement of broken units, to minimize delays caused by waiting for repairs.  When the sites database was 'cleaned' at the end of the survey stage, serious errors (ie. fields containing data outside of a credible range or containing a mix of integers and non-integers, mostly likely due to accidents during data entry) and omissions (fields left blank or with data that is incomplete) were found to have occurred in fewer than 0.5% of the sites. Sites with problematic data were re-recorded and the unclean recorded were marked so they could be excluded from analysis.

@@

Sullivan, M., P. Hiscock, and P. Hughes (2014) Three Scales: GIS, GPS and Digital Site and Data Recording Technology in Archaeological Salvage at Olympic Dam in Arid South Australia. Journal of the Anthropological Society of South Australia 38: 85-107.
